---
title: "hw04-gapminder"
author: '@RetoSterchi'
date: "September 30, 2018"
output: github_document
---

# Homework 04: Tidy data and joins

## First step: Load packages

```{r message=FALSE, warning=FALSE}
require(gapminder)
require(ggplot2)
require(dplyr)
require(tidyr)
```


## Data Reshaping Activity #2
### Task description

Make a tibble with one row per year and columns for life expectancy for two or more countries.

 - Use `knitr::kable()` to make this table look pretty in your rendered homework.
 - Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

Let's use the *spread* function to re-arrange our data into a wide format.  (https://tidyr.tidyverse.org/reference/spread.html)
```{r}
knitr::kable(
  
  gapminder %>%
  filter(country %in% c("Canada", "Japan", "Rwanda")) %>% 
  select(year, country, lifeExp) %>% ## select columns 
  spread(country, lifeExp) %>% ## reshape to wide
  rename(Year = year) ## rename column for table

)
```

Now let's take this re-shaped data and make a scatterplot that shows life expectancy for one country against that of another.

```{r}

gapminder %>%
  
  ## Reshape data (same as above)
  filter(country %in% c("Canada", "Japan", "Rwanda")) %>% 
  select(year, country, lifeExp) %>% ## select columns 
  spread(country, lifeExp) %>% ## reshape to wide
  rename(Year = year) %>% ## rename column for table
  
  ## plot the data
  ggplot(aes(x = Rwanda, y = Canada)) + ## set up plot
  geom_point() + ## ad points
  xlim(35,85) + ## use custom x scale
  ylim(35,85) + ## use custom y scale
  coord_fixed() + ## makes x/y axes equal scaled, squared
  geom_abline(intercept = 0, slope = 1) + ## add diagonal line for reference
  ggtitle("Life expectancy Canada vs. Rwanda") ## add title
  

```

Looks like the life expectancy in Canada is almsost twice as high as the one from Rwanda.

## Data Join Activity #1

### Task description
Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a `dplyr` join function and make some observations about the process and result. Explore the different types of joins. Examples of a second data frame you could build:

 - One row per country, a country variable and one or more variables with extra info, such as language spoken, NATO membership, national animal, or capitol city.
 - One row per continent, a continent variable and one or more variables with extra info, such as northern versus southern hemisphere.

Let's create that supplementray data.frame first with some information in it.

```{r}
df <- data.frame("country" = c("Canada", "Japan", "Sweden", "Mexico", "Russia", "France"),
                 "language" = c("English", "Japanese", "Swedish", "Spanish", "Russian", "French"),
                 "hemisphere" = c("N", "N", "N", "N", "N", "N"),
                 "capitalcity" = c("Ottawa", "Tokyo", "Stockholm", "Mexico City", "Moscow", "Paris"))

## when joining the dataframe later on with gapminder we received the following warning message:
## ...
## Joining, by = "country"
## Column `country` joining factors with different levels, coercing to character vector
## ...

## Although it seems to work, let's make sure that we have the data already in the correct format and 
## let's assign the proper factor levels to our data.frame

df$country <- factor(df$country, levels = levels(gapminder$country))
```

There are different types of joins (https://dplyr.tidyverse.org/reference/join.html)
Mutating joins combine variables from the two data.frames:

 - `inner_join()` return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.
 - `left_join()` return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.
 - `right_join()` return all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.
 - `full_join()` return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.

```{r}
gapminder %>% inner_join(df, by = "country")
gapminder %>% left_join(df, by = "country")
gapminder %>% right_join(df, by = "country")
gapminder %>% full_join(df, by = "country")
```

This corresponds to the following merges in base R:

```{r eval=F}
merge(gapminder, df, by = "country") ## merge to matches in both sides
merge(gapminder, df, by = "country", all.x = TRUE) ## merge to matches on the left
merge(gapminder, df, by = "country", all.y = TRUE) ## merge to matches on the right
merge(gapminder, df, by = "country", all = TRUE) ## merge both sides, fill NA's if no matches
```